---
title: "dplyr-basics"
format: html
---

## SELECT

## FILTER

## MUTATE

## JOIN (inner, left,right,anti)

# Diamond Data Set

```{r}
library(dplyr)
library(ggplot2)
data(diamonds)

head(diamonds)

# selct by name

sub_set <- select(diamonds,c("carat","cut"))
sub_set <- select(diamonds,carat,cut)
sub_set <- select(diamonds,carat:cut)
sub_set <- select(diamonds,-c(1,4)) # exclude
head(sub_set)

# select column using helper functions

diamondC <- select(diamonds, starts_with("C"))
head(diamondC)
diamondR <- select(diamonds, ends_with("R"))
head(diamondR)

diamonds <- rename(diamonds, Quality = cut)
head(diamonds)

sample_n(diamonds,20)


# create dataframe
df <- data.frame(player=c('A','B','C','D','E','F','G','H','I'),
    points=c(4,7,8,12,14,16,20,26,36)
)


df$category <- cut(df$points,breaks=4)

df$category <- cut(df$points,breaks=c(0,10,15,20,40))
df$category <- cut(df$points,breaks=c(0,10,15,20,40),
labels=c('Bad','OK','Good','Great')
)

# find the mean price for each of the following weight categories:
# 0,0.25,0.5,0.75,1,100
diamonds$category3 <- cut(diamonds$carat,breaks=c(0,0.25,0.5,0.75,1,100))
# create frequency table
freq_table <- table(diamonds$category3)
print(freq_table)

hist(freq_table)

```

```{r}
# Load dplyr to use the data
library(dplyr)

# 1. The Primary Dataset (Used for the 5 basic verbs)
df <- data.frame(
  ID = 1:10,
  Name = c("Alice", "Bob", "Charlie", "David", "Eve", "Frank", "Grace", "Heidi", "Ivan", "Judy"),
  Age = c(28, 35, 32, 45, 29, 50, 31, 41, 27, 38),
  Department = c("IT", "Sales", "IT", "HR", "Marketing", "IT", "Sales", "HR", "IT", "Marketing"),
  Salary = c(75000, 60000, 82000, 90000, 65000, 105000, 72000, 85000, 88000, 74000)
)


```

## Select, filter, arrange,summary

```{r}

clean_df <- select(df,ID,Age,Department) # select column
filter_df <- filter(df,(Salary>=105000 & Department=='IT') | Name=='Charlie')

mutate_df <- mutate(df,Bonus=Salary*0.10)

aranged_df <- arrange(df,desc(Salary),Department)
aranged_df
# alternate way
clean_df <- df %>% select(ID,Age,Department)

dept_summary <- 
  group_by(df,Department) |> 
  summarize(
    Total_Payroll = sum(Salary, na.rm = TRUE),
    Average_Age = mean(Age, na.rm = TRUE)
  )


```

```{r}
# 2. Table A for Joins (Employees)
employees <- data.frame(
  emp_id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Charlie", "David"),
  dept_id = c("D1", "D2", "D1", "D99")
)

# 3. Table B for Joins (Departments)
departments <- data.frame(
  dept_id = c("D1", "D2", "D3"),
  dept_name = c("Engineering", "Sales", "HR")
)

joined <- inner_join(employees,departments,by="dept_id")
joined
left_joined <- left_join(employees,departments,by=c("dept_id"="dept_id"))
left_joined
```

```{r}
library(dplyr)

# Global Engineering Infrastructure Log
set.seed(42)
infra_pipeline_df <- data.frame(
  log_id = 101:120,
  timestamp = seq(from = as.POSIXct("2026-05-27 00:00:00"), by = "30 mins", length.out = 20),
  node_cluster = c("Cluster_Alpha", "Cluster_Beta", "Cluster_Alpha", "Cluster_Gamma", "Cluster_Beta", 
                   "Cluster_Alpha", "Cluster_Gamma", "Cluster_Beta", "Cluster_Alpha", "Cluster_Alpha",
                   "Cluster_Beta", "Cluster_Beta", "Cluster_Alpha", "Cluster_Gamma", "Cluster_Gamma",
                   "Cluster_Alpha", "Cluster_Beta", "Cluster_Gamma", "Cluster_Alpha", "Cluster_Beta"),
  aws_service = c("Glue", "Kinesis", "S3", "Redshift", "SageMaker", "Glue", "S3", "Kinesis", "Redshift", "Glue",
                  "SageMaker", "S3", "Glue", "Redshift", "Kinesis", "SageMaker", "Glue", "S3", "Redshift", "Kinesis"),
  payload_size_kb = c(1420, 2800, 50, 9400, 15200, 1100, 45, 3100, NA, 1950, 
                      21000, 120, 850, 11500, 4100, 18400, NA, 85, 13200, 3900),
  execution_time_sec = c("0.45", "1.20", "0.05", "5.40", "12.80", "0.38", "0.04", "1.45", "7.20", "0.52",
                         "18.10", "0.11", "TIMEOUT", "6.10", "1.95", "14.20", "0.55", "0.08", "8.95", "1.60"),
  status_code = c(200, 200, 200, 500, 200, 200, 200, 200, 502, 200, 
                  500, 200, 504, 200, 200, 200, 200, 404, 200, 200),
  is_retry = c(FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, TRUE, TRUE, FALSE,
               TRUE, FALSE, TRUE, FALSE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE),
  stringsAsFactors = FALSE
)

head(infra_pipeline_df)
```

### Question 1: The Outlier Network Scan

Filter the dataset down to rows belonging to **Cluster_Alpha** or **Cluster_Gamma** where the `payload_size_kb` is strictly *above* the global median payload size of the entire dataset. However, your filter must safely handle missing values: if a payload size is `NA`, it must be preserved and included in the final output *only* if the `is_retry` flag is `TRUE`.

```{r}
filtered_data <- infra_pipeline_df |> 
  filter(
    # 1. Lock down the targeted clusters
    node_cluster %in% c("Cluster_Alpha", "Cluster_Gamma"),
    
    # 2. Apply the conditional payload logic safely handling NAs
    (payload_size_kb > median(payload_size_kb, na.rm = TRUE)) | (is.na(payload_size_kb) & is_retry == TRUE)
  )

filtered_data

```

### Question 2: The Multi-Step Relational Intersect

Without modifying the original data frame, isolate all records belonging to any `aws_service` whose *average* payload size (calculated across the whole dataset, ignoring `NA`s) is strictly greater than $3,000\text{ KB}$.

```{r}
average_payload <- mean(infra_pipeline_df$payload_size_kb)
aws_services <- c('Glue','Kinesis','Redshift')
aws_data <- infra_pipeline_df |> filter(payload_size_kb>300 & aws_service %in% aws_services & timestamp> as.POSIXct("2026-05-27"))
aws_data
```

```{r}
imputed_df <- infra_pipeline_df |> mutate(payload_size_kb,coalesce(payload_size_kb,0))
imputed_df

```

```{r}
library(dplyr)

cluster_nodes <- data.frame(
  node_id = c("N_101", "N_102", "N_103", "N_104"),
  cluster_name = c("Alpha", "Alpha", "Beta", "Gamma"),
  region = c("us-east-1", "us-east-1", "us-west-2", "eu-central-1")
)

pod_metrics <- data.frame(
  pod_id =  c("P_01", "P_02", "P_03", "P_04", "P_05", "P_06"),
  node_id = c("N_101", "N_101", "N_102", "N_103", "N_103", "N_999"), # N_999 is an orphaned pod
  cpu_utilization = c(85.5, 92.1, 45.0, 78.2, 88.9, 12.0),
  memory_gb = c(16, 32, 8, 16, 64, 4)
)
```

### Quesion 1: Convert this SQL to equivalent dplyr

```{sql connection=}
SELECT 
    n.region,
    SUM(p.memory_gb) AS total_memory_allocated,
    COUNT(p.pod_id) AS total_active_pods
FROM pod_metrics p
INNER JOIN cluster_nodes n ON p.node_id = n.node_id
WHERE p.cpu_utilization > 50.0 
  AND n.region != 'eu-central-1'
GROUP BY n.region
HAVING COUNT(p.pod_id) >= 1
ORDER BY total_memory_allocated DESC
LIMIT 2 OFFSET 1;
```

```{r}
data <- inner_join(pod_metrics, cluster_nodes, by = "node_id") |> 
  filter(cpu_utilization > 50 & region != 'eu-central-1') |> 
  group_by(region) |>
  summarize(
    total_memory_allocated = sum(memory_gb),
    total_active_pods = n()  # Changed to n()
  ) |> 
  arrange(desc(total_memory_allocated)) |> # Fixed typo to desc()
  slice(2:3)                                # Added to handle LIMIT & OFFSET

data

```
