---
title: "Data Manipulation Mastery with Dplyr"
author: "Analysis Notebook"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(readr)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 1. Introduction to the Dataset

We begin by loading the built-in `diamonds` dataset and inspecting its structure. This dataset contains the prices and other attributes of almost 54,000 diamonds.

```{r exploration}
data(diamonds)

# Quick look at the first few rows
head(diamonds)

# Check the data types (Structure)
str(diamonds)
```

------------------------------------------------------------------------

# 2. Selecting and Dropping Columns

The `select()` function allows us to focus on specific variables or remove those we don't need.

### Selecting by Name or Range

```{r selection}
# Select specific columns and ranges
subset_diamonds <- select(diamonds, carat, color:clarity)

# Select columns using helper functions
diamondC <- select(diamonds, starts_with("C"))
diamondR <- select(diamonds, ends_with("R"))
```

### Dropping Columns

```{r dropping}
# Remove table and price by name
diamondDrop <- select(diamonds, -c(table, price))

# Remove by index (Column 1 and 5)
diamondDrop_idx <- select(diamonds, -c(1, 5))
```

------------------------------------------------------------------------

# 3. Filtering and Sorting

Filtering allows us to zoom in on specific rows based on logical conditions.

```{r filtering}
# Filter for large diamonds (>1 carat)
diamond1 <- filter(diamonds, carat > 1)

# Filter for Premium cut diamonds
diamondPremium <- filter(diamonds, cut == "Premium")

# Complex filtering (Carat and Color)
diamondsHigh <- filter(diamonds, carat > 0.75 & color == "D")
```

### Sorting Data

Use `arrange()` to order your data. By default, it sorts in ascending order.

```{r sorting}
diamondSort <- arrange(diamonds, price)
# View cheapest and most expensive
head(diamondSort, 3)
tail(diamondSort, 3)
```

------------------------------------------------------------------------

# 4. Mutating and Renaming

Adding new information or cleaning up column names.

```{r transformation}
# Rename 'cut' to 'Quality'
diamondRN <- rename(diamonds, Quality = cut)

# Calculate how far each diamond is from the average price
diamondDiff <- mutate(diamonds, PRICEdiff = price - mean(price, na.rm = TRUE))

# Get a random sample to check results
sample_n(diamondDiff, size = 4)
```

------------------------------------------------------------------------

# 5. Grouping and Summarization

This is the "Split-Apply-Combine" strategy for data analysis.

```{r summary}
# Group by clarity and calculate statistics
CLARITY_summary <- diamonds %>%
  group_by(clarity) %>%
  summarize(
    PRICE.Average = mean(price, na.rm = TRUE), 
    PRICE.Max = max(price, na.rm = TRUE), # Fixed from min to max for logic
    PRICE.Median = median(price, na.rm = TRUE)
  )

print(CLARITY_summary)
```

------------------------------------------------------------------------

# 6. Categorization (Binning)

Using `cut()` to turn continuous numbers into discrete categories.

```{r binning}
# Create custom weight categories for carats
diamonds$category3 <- cut(diamonds$carat, 
                          breaks = c(0, 0.25, 0.5, 0.75, 1, 100),
                          labels = c("Very Small", "Small", "Medium", "Large", "Huge"))

# Analyze average price by these new categories
category_summary <- diamonds %>%
  group_by(category3) %>%
  summarize(PRICE.Average = mean(price, na.rm = TRUE))

print(category_summary)
```

------------------------------------------------------------------------

# 7. Joining Data

When working with multiple files, we use joins to merge them based on a common ID.

> **Note:** The following code assumes the files exist at the specified path.

```{r joins, eval=FALSE}
test1 <- read_csv("/home/tekraj/MSc-Mathematics-Github/test1.csv")
test2 <- read_csv("/home/tekraj/MSc-Mathematics-Github/test2.csv")

# Combine datasets using an inner join
combined_data <- inner_join(test1, test2, by = "id")
```

------------------------------------------------------------------------

# 8. Visualizing the Results

Finally, let's look at the relationship between Carat and Price using `ggplot2`.

```{r plot}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point(alpha = 0.3) +
  theme_minimal() +
  labs(title = "Diamond Price vs Carat",
       subtitle = "Colored by Cut Quality",
       x = "Weight (Carat)",
       y = "Price ($)")
```
