---
title: "R Notebook"
output: html_notebook
---

# GGPlot Basics

### Load Iris Data

create a scatter plot to see if there is a relationship between the length of a flower's sepal and its petal, color-coded by which species it belongs to.

```{r}
library(ggplot2)
data(iris)

irisPlot <- ggplot(iris, aes(Sepal.Length, Petal.Length, colour = Species))+ geom_point()
print(irisPlot)
print(irisPlot+labs(y="Petal Length(cm)",x="Sepal Length(cm)")) + ggtitle("Petal and sepal length of iris")
# annotate
irisPlot + annotate("text", x=6, y=5, label="text")
# highlight an area
irisPlot+annotate("rect",xmin = 5, xmax = 7, ymin=4, ymax=6, alpha=0.5)

# segment 
irisPlot + annotate("segment", x=5, xend=7, y = 4, yend=5, color="black")
```

```{r}
df <- data.frame(x = c(1, 2, 3, 4, 5), 
                 y = c(10, 14, 18, 24, 30))

ggplot(data = df, aes(x = x, y = y,color=x)) + 
  geom_point()

```

## Common "Geoms" for two variables

Once we have the first line set up, you can swap out `geom_point()` for other types of charts:

|                  |                  |
|------------------|------------------|
| **Goal**         | **Layer to add** |
| **Scatter Plot** | `+ geom_point()` |
| **Line Graph**   | `+ geom_line()`  |
| **Bar Chart**    | `+ geom_col()`   |
| **Area Chart**   | `+ geom_area()`  |

# Plant Growth Data

```{r}
data(PlantGrowth)
head(PlantGrowth)
bp <- ggplot(PlantGrowth, aes(x=group, y = weight))+geom_point()
print(bp)

# swipe x and y as per our requirements

bp <- ggplot(PlantGrowth, aes(x=weight, y = group)) + geom_point()
print(bp)
```

# Multi Panel Plots

```{r}
library(tidyverse)
data(airquality)
head(airquality)
temperature <- airquality$Temp
ozone <- airquality$Ozone
par(mflow=c(2,2))
hist(temperature)
boxplot(temperature,horizonal=TRUE)
hist(ozone)
```

# Ggplot 2- Marginal Plots

``` {r}
 library(ggplot2)
 library(ggExtra)
 data("mpg")
 head(mpg)
 g <- ggplot(mpg,aes(cty,hwy))+geom_count() + geom_smooth(method="lm",se=F)
 print(g)
 ggMarginal(g, type="boxplot", fill="transparent")
 ggMarginal(g,type="histogram",fill="transparent")
 
```

## Multiple Plots

```{r}

p <- ggplot(mpg,aes(cty)) + geom_density(aes(fill=factor(cyl)),alpha=0.8)
print(p)

p+ labs(title="Density Plot", subtitle="City Mileage Grouped by Number of Cylinders"
        ,caption = "Source: mpg", x="City Mileage", fill="#Cylinders")
```