---
title: "R-Basics"
format: html
---

## R Data Types

-   Numeric (float)

-   Integer

-   Character(String)

-   Logical(Boolean)

-   Complex

```{r}
a <- 5.25 # numeric (float)
print(class(a))
b <- 5L # integer
print(class(b))


comp <- 1+3i # complex
print(class(comp))

name <- "Ram" # character
print(class(name))

pass = FALSE # logical
print(class(pass))


```

### Utility Commands to Know related to Data Types

-   **`class(x)`**: Tells us the data type of the variable.

<!-- -->

-   **`typeof(x)`**: Reveals the internal, low-level storage type in R (e.g., `typeof(42)` will return `"double"` instead of `"numeric"`).

-   **`is.*` functions**: Used to check if a variable matches a specific type, returning `TRUE` or `FALSE`.

-   as.data type: Used to cast data types

```{r}
a <- 5
print(class(a))
print(typeof(a))
print(is.numeric(a)) # true if a is numeric
print(is.integer(a))
print(as.character(a)) # converts to character

```

## R Basic Arithematics

-   **+** : Addition

<!-- -->

-   **-** : Subtraction

-   \* : Multiplication

-   **/** : Division

-   **\^ or** : Exponentiation (Power)

-   **%%** : Modulo (Remainder)

-   **%/%** : Integer Division

```{r}
a <- 5
b <-7

print("SUM")
sum <- a+b
print(sum)

print("Subtraction")
sub <- a-b
print(a-b)

print("Multiply")
prod <- a*b
print(prod)

print("divide")
div <- a/b
print(div)

print("power")
power <- a^b
pw <- a**b
print(power)
print(pw)

print("Reminder")
rem <- a%%b
print(rem)


print("Int Division")
int_div <- a%/%b
print(int_div)
```

## Built-in Mathematical Functions

### Rounding & Absolute Values

-   **`abs(x)`**: Absolute value (converts negative numbers to positive).

-   **`sign(x)`**: Returns `-1` for negative numbers, `0` for zero, and `1` for positive numbers.

-   **`sqrt(x)`**: Square root ($\sqrt{x}$).

-   **`round(x, digits = n)`**: Rounds a number to `n` decimal places.

-   **`ceiling(x)`**: Rounds *up* to the nearest integer (e.g., `ceiling(3.1)` returns `4`).

-   **`floor(x)`**: Rounds *down* to the nearest integer (e.g., `floor(3.9)` returns `3`).

-   **`trunc(x)`**: Truncates the decimal part entirely, leaving just the integer (behaves like `floor` for positive numbers and `ceiling` for negative numbers).

    ```{r}
    a <- 100
    b<- -10.46
    print("abs")
    print(abs(b))

    print("sqrt")
    print(sqrt(a))

    print("round")
    print(round(b,1))


    print("ceiling")
    ceiling(b)


    print('floor')
    floor(b)

    print('truncate')
    trunc(b)

    ```

### Logarithms & Exponentials

-   **`exp(x)`**: Exponential function ($e^x$).

-   **`log(x)`**: Natural logarithm ($\ln(x)$ with base $e$).

-   **`log10(x)`**: Base-10 logarithm ($\log_{10}(x)$).

-   **`log2(x)`**: Base-2 logarithm ($\log_2(x)$).

-   **`log(x, base = b)`**: Logarithm with a custom base $b$.

    ```{r}
    a <- 100

    print('exp')
    print(exp(a))

    print("log base e")
    print(log(a))

    print('log10')
    print(log10(a))

    print(log(a,base=2))
    ```

## Trigonometric Functions

R expects all angles for trigonometric functions to be in **radians**, not degrees.

-   **`sin(x)` / `cos(x)` / `tan(x)`**: Sine, cosine, and tangent.

-   **`asin(x)` / `acos(x)` / `atan(x)`**: Inverse trig functions (arcsine, arccosine, arctangent).

-   **`sinh(x)` / `cosh(x)` / `tanh(x)`**: Hyperbolic functions.

```{r}
angle <-45
rad <- angle * (pi/ 180)

print(tan(rad))


```

## Built in constants

-   **`pi`** : The mathematical constant $\pi$ (3.141593...)

-   **`letters`** : The 26 lowercase English alphabet letters ("a" to "z")

-   **`LETTERS`** : The 26 uppercase English alphabet letters ("A" to "Z")

-   **`month.name`** : The full names of the 12 months ("January" to "December")

-   **`month.abb`** : The 3-letter abbreviations for the 12 months ("Jan" to "Dec")

-   **`NA`** : Represents missing data ("Not Available")

-   **`NaN`** : Represents undefined numerical results ("Not a Number")

-   **`Inf`** : Positive infinity

-   **`-Inf`** : Negative infinity

```{r}
month.name
letters
```
