---
title: "NEPSE-EDA"
format: html
editor: visual
---

# Banking Sector Data

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(moments)
library(boot)
library(WRS2)
library(lmPerm)
library(tidyr)
library(purrr)
```

```{r}
nepse_data <- read.csv("./data/nepse.csv")
# sort data in ascending order by Date
nepse_data <- nepse_data |>
  mutate(Date = as.Date(Date)) |>
  arrange(Date,)

head(nepse_data,5)
```

## Exploratory Data Analysis

### Box Plot

```{r}

# box plots
ggplot(nepse_data, aes(x = Symbol, y = Close, fill = Symbol)) +
  geom_boxplot() +
  labs(title = "Box Plot of Close Price by Symbol", 
       x = "Symbol", 
       y = "Close Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Histogram

```{r}
ggplot(nepse_data, aes(x = Close, fill = Symbol)) +
  geom_histogram(bins = 30, color = "white") +
  facet_wrap(~Symbol, scales = "free") + # 'scales = "free"' allows different axes per category
  labs(title = "Histogram of Close Price by Symbol", 
       x = "Close Price", 
       y = "Frequency") +
  theme_minimal()
```

### Q-Q Plot

```{r}
ggplot(nepse_data, aes(sample = Close, color = Symbol)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~Symbol, scales = "free") +
  labs(title = "Q-Q Plot of Close Price by Symbol",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

## Probability Distribution or Return

```{r}
# Calculate daily percentage returns
sector_returns <- nepse_data |>
  group_by(Symbol) |>
  arrange(Date) |>
  mutate(Return = (Close - lag(Close)) / lag(Close)) |>
  filter(!is.na(Return)) # Remove the first row which becomes NA

# Summarize distribution metrics
distribution_stats <- sector_returns |>
  summarise(
    Mean = mean(Return),
    SD = sd(Return),
    Skewness = skewness(Return),
    Kurtosis = kurtosis(Return)
  )

print(distribution_stats)
```

### Normality Testing Using Shapiro-Wilk Test

```{r}
normality_sw <- sector_returns %>%
  group_by(Symbol) %>%
  summarise(p_value = shapiro.test(Return)$p.value)

print(normality_sw)
```

## Mathematical Modeling

### Simple Linear Regression

For simple linear trends (time-based) and non-linear patterns (using polynomials), we can fit these models:

```{r}
# Simple Linear: Return ~ Date
linear_model <- lm(Return ~ as.numeric(Date), data = sector_returns)

# Non-linear (Quadratic): Return ~ Date + Date^2
nonlinear_model <- lm(Return ~ poly(as.numeric(Date), 2), data = sector_returns)

summary(linear_model)
```

### Multiple Linear Regression

We will predict the `Return` of a specific sector using its own `lag(Return)` and the `Market_Index` (if you have one; here I demonstrate using `lagged` returns as an autoregressive model).

```{r}
# Prepare lagged data for regression
model_data <- sector_returns %>%
  group_by(Symbol) %>%
  mutate(Lag_Return = lag(Return)) %>%
  filter(!is.na(Lag_Return))

# Regression: Return vs Lagged Return
multi_reg <- lm(Return ~ Lag_Return + Symbol, data = model_data)
summary(multi_reg)
```

### Chi-Square Test

To study the dependence between return direction (Gain/Loss) and the Month:

```{r}
chi_data <- sector_returns %>%
  mutate(Direction = ifelse(Return > 0, "Gain", "Loss"),
         Month = month(Date, label = TRUE))

# Create Contingency Table
ctable <- table(chi_data$Month, chi_data$Direction)

# Perform Chi-Square Test
chisq_test <- chisq.test(ctable)
print(chisq_test)
```

### Analysis of Variance (ANOVA)

ANOVA tests if the mean returns differ significantly across months or sectors.

```{r}
# ANOVA: Return by Month
anova_month <- aov(Return ~ factor(month(Date)), data = sector_returns)
summary(anova_month)

# ANOVA: Return by Sector
anova_sector <- aov(Return ~ Symbol, data = sector_returns)
summary(anova_sector)
```

### Non-Parametric Tests

If your normality tests (from the previous step) failed, use these robust alternatives:

```{r}
# Kruskal-Wallis (Comparing Returns across Symbols)
kw_test <- kruskal.test(Return ~ Symbol, data = sector_returns)
print(kw_test)

# Mann-Whitney (Comparing two specific sectors, e.g., BANKING vs OTHERS)
# Subset data for two groups
subset_data <- sector_returns %>% filter(Symbol %in% c("BANKING", "OTHERS"))
mw_test <- wilcox.test(Return ~ Symbol, data = subset_data)
print(mw_test)
```

## Seasonal Trading Strategies

### 1. Monthly Ranking – Best & Worst Months

```{r}
monthly_returns <- sector_returns %>%
  mutate(Month = month(Date, label = TRUE)) %>%
  group_by(Symbol, Month) %>%
  summarise(
    Mean_Return   = mean(Return, na.rm = TRUE),
    Median_Return = median(Return, na.rm = TRUE),
    CI_Lower      = quantile(Return, 0.025, na.rm = TRUE),
    CI_Upper      = quantile(Return, 0.975, na.rm = TRUE),
    Win_Ratio     = sum(Return > 0) / n(),
    Sharpe        = mean(Return) / sd(Return),
    .groups       = "drop"
  ) %>%
  arrange(Symbol, Month)

best_months <- monthly_returns %>%
  group_by(Symbol) %>%
  slice_max(Mean_Return, n = 2) %>%
  ungroup()

cat("\n===== Best months per sector =====\n")
print(best_months)

worst_months <- monthly_returns %>%
  group_by(Symbol) %>%
  slice_min(Mean_Return, n = 2) %>%
  ungroup()

cat("\n===== Worst months per sector =====\n")
print(worst_months)
```

### 2. Z-Score Entry/Exit Framework

```{r}
zscore_analysis <- sector_returns %>%
  mutate(Month = month(Date, label = TRUE)) %>%
  group_by(Symbol, Month) %>%
  mutate(
    Monthly_Mean = mean(Return, na.rm = TRUE),
    Monthly_SD   = sd(Return, na.rm = TRUE),
    Z_Score      = (Return - Monthly_Mean) / Monthly_SD
  ) %>%
  ungroup()

signals <- zscore_analysis %>%
  filter(abs(Z_Score) > 1.5) %>%
  mutate(Signal = ifelse(Z_Score < -1.5, "ENTRY", "EXIT")) %>%
  group_by(Symbol, Signal, Month) %>%
  summarise(
    Frequency = n(),
    Avg_Z     = mean(Z_Score),
    .groups   = "drop"
  )

cat("\n===== Z-score signals (|Z|>1.5) =====\n")
print(signals)
```

### 3. Cumulative Return by Day of Year

```{r}
cumulative_pattern <- sector_returns %>%
  mutate(DayOfYear = yday(Date)) %>%
  group_by(Symbol, DayOfYear) %>%
  summarise(Mean_Return = mean(Return, na.rm = TRUE), .groups = "drop") %>%
  group_by(Symbol) %>%
  mutate(Cumulative_Mean = cumsum(Mean_Return))

optimal_exit <- cumulative_pattern %>%
  group_by(Symbol) %>%
  slice_max(Cumulative_Mean, n = 1, with_ties = FALSE) %>%
  mutate(Exit_Day = DayOfYear,
         Exit_Date_approx = as.Date(paste0("2024-", Exit_Day), format = "%Y-%j"))

cat("\n===== Optimal exit day-of-year per sector =====\n")
print(optimal_exit %>% select(Symbol, Exit_Day, Cumulative_Mean))

p_cum <- ggplot(cumulative_pattern, aes(x = DayOfYear, y = Cumulative_Mean, color = Symbol)) +
  geom_line() +
  labs(title = "Cumulative mean return by day of year",
       x = "Day of year", y = "Cumulative return") +
  theme_minimal()
print(p_cum)

ggplot(cumulative_pattern, aes(x = DayOfYear, y = Cumulative_Mean, color = Symbol)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~Symbol, scales = "free_y") +
  labs(title = "Cumulative Mean Returns by Day of Year",
       subtitle = "Comparing performance trends across NEPSE sectors",
       x = "Day of Year (1-366)",
       y = "Cumulative Return") +
  theme_minimal() +
  theme(legend.position = "none")
```

### 4. Year-by-Year Consistency (≥80% Positive Years)

```{r}
consistency_test <- sector_returns %>%
  mutate(Year = year(Date), Month = month(Date, label = TRUE)) %>%
  group_by(Symbol, Month, Year) %>%
  summarise(Monthly_Return = sum(Return, na.rm = TRUE), .groups = "drop") %>%
  group_by(Symbol, Month) %>%
  summarise(
    Positive_Years   = sum(Monthly_Return > 0),
    Total_Years      = n(),
    Consistency_Rate = Positive_Years / Total_Years,
    .groups = "drop"
  ) %>%
  filter(Total_Years >= 3)

reliable_seasons <- consistency_test %>% filter(Consistency_Rate >= 0.8)

cat("\n===== Reliable seasonal months (≥80% positive years) =====\n")
print(reliable_seasons)
```

### 5. Bootstrap Confidence Intervals (All Sector-Month Pairs)

```{r}
sector_months <- sector_returns %>%
  distinct(Symbol, Month = month(Date)) %>%
  mutate(Returns_list = map2(Symbol, Month, ~ {
    sector_returns %>%
      filter(Symbol == .x, month(Date) == .y) %>%
      pull(Return)
  }))

boot_mean <- function(x, indices) {
  mean(x[indices], na.rm = TRUE)
}

bootstrap_results <- sector_months %>%
  mutate(
    Boot = map(Returns_list, ~ {
      if (length(.x) < 2) return(tibble(CI_lower = NA, CI_upper = NA))
      set.seed(123)
      b <- boot(.x, boot_mean, R = 1000)
      ci <- tryCatch(boot.ci(b, type = "perc")$percent[4:5],
                     error = function(e) c(NA, NA))
      tibble(CI_lower = ci[1], CI_upper = ci[2])
    })
  ) %>%
  unnest(Boot) %>%
  select(-Returns_list)

cat("\n===== Bootstrap percentile CIs for all sector-month pairs =====\n")
print(head(bootstrap_results, 12))
```

### 5b. Bootstrap CI – BANKING in Kartik (Month 8)

```{r}
banking_kartik_returns <- sector_returns %>%
  filter(Symbol == "BANKING", month(Date) == 8) %>%
  pull(Return)

if (length(banking_kartik_returns) >= 2) {
  set.seed(123)
  boot_bank_kart <- boot(banking_kartik_returns, boot_mean, R = 5000)
  ci_bank_kart <- boot.ci(boot_bank_kart, type = "perc")
  cat("\n===== Specific CI for BANKING in Kartik (month 8) =====\n")
  print(ci_bank_kart)
}
```

### 6. Robust ANOVA (Trimmed Means) and Permutation Test

```{r}
robust_anova_month <- t1way(Return ~ factor(month(Date)), data = sector_returns, tr = 0.2)
cat("\n===== Robust ANOVA (20% trimmed means) – Month effect =====\n")
print(robust_anova_month)

set.seed(123)
perm_anova_month <- aovp(Return ~ factor(month(Date)), data = sector_returns,
                         np = 1000)
cat("\n===== Permutation ANOVA – Month effect =====\n")
summary(perm_anova_month)
```

### 7. Two-Way Interaction ANOVA: Symbol × Month

```{r}
sector_returns <- sector_returns %>%
  mutate(MonthFactor = factor(month(Date)))

interaction_anova <- aov(Return ~ Symbol * MonthFactor, data = sector_returns)
cat("\n===== Two-way ANOVA (Symbol × Month) =====\n")
summary(interaction_anova)
```

### 8. Monthly Means with Confidence Intervals

```{r}
print(head(monthly_returns %>% select(Symbol, Month, Mean_Return, CI_Lower, CI_Upper, Win_Ratio), 12))
```

### 
